Leaf traits drive differences in biomass partitioning among major plant functional types

### 2015-01-13 manuscript.Rnw
Added R2 for LMA instead of PFT in models of LMF and MF/AS. Editing to text; added comment to paragraph about AF/AS.

The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\cite{ise2010}. Globally, forests store approximately 360Pg of carbon\citep{panetal2011} in living biomass, equivalent to about 40years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}. Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass determines turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}. Reducing uncertainties about biomass partitioning and carbon residence times in GVMs is therefore a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014}.

@@

Moreover, we show that differences in LMF among woody PFTs arise due to differences in leaf morphology, but they converge in the amount of leaf area held on the plant (Fig. 2b-c). Previously, it was unclear whether PFTs with higher leaf biomass also had higher total leaf area. Leaf area is relevant because it drives total light interception and thus photosynthesis. Several classic studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}. We found no evidence for this, instead we found that the three PFTS maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR). Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig. 2a and S2). We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}. We then found that LMF at a common height was proportional to the leaf construction cost, as measured by leaf mass per area (LMA) (Fig. 2b). As a consequence, LAR was invariant between PFTs, because $\textrm{LAR} = LMF / \textrm{LMA}$ (Fig. 2c and S3).

Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem ($A_S$), a quantity closely related to hydraulic supply capacity \cite{enquist2003}. This ratio is also important because plant leaf mass is frequently estimated from $A_S$ with allometric equations \cite{shinozaki1964b, chave2005}.  Mirroring the results for LMF, we find that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig. 2d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig. 2e).

The above results highlight how differences in LMA -- a central trait of the leaf economics spectrum\citep{wrightetal2004} -- drives differences in biomass partitioning among dominant woody PFTs. Based on a XXX statistical model, we found that LMA alone could explain almost as much  (?more?) variation as PFT, even though LMA only captures aspects of leaf morphology. These results suggest promising avenues for parameterizing and simplifying GVMs.

@@

In contrast, we found that variation among species in leaf and stem traits could explain additional variation in biomass partitioning. The effect of LMA was found to hold also within PFTs (Fig. 4a), explaining 28\% of the variation in $M_F/A_S$ across species. The fact that PFTs differ in LMA is evident in the separation of the three groups along the x-axis of Fig. 4a. Yet these clouds were broad and overlapping, with a more general relationship extending across the entire LMA axis. The remaining (?? what is number??) variation in $M_F/A_S$  was explained by species-level differences in $A_F/A_S$ (Fig. 4b). As previously noted, $A_F/A_S$ does not differ systematically among PFTs (Fig. 2e), but does vary close to two orders of magnitude across species. The separation among PFTs along the Y-axis arises due to differences in LMA.

@@

We used the recently compiled Biomass And Allometry Database (BAAD) \citep{Falster2015}, including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_W$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca. 1.3m) ($A_{Sbh}$).  For each analysis, we used different subsets of the data. Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4} We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).

@@

We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves). For conifers, leaf area was converted to half-total surface area using published conversion factors \citep{barclay2000}, which is most appropriate for comparison to flat leaves in terms of light interception  \citep{lang1991, chen1992}. Stem cross-sectional area was measured either at breast height (ca. 1.3m) and/or at the base of the plant. For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using non-linear regression (see~\ref{sec:predas}).

@@

Figure 1a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary. We thank Martin de Kauwe for discussions and commenting on an earlier version of this paper.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
    \caption{\textbf{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_W$) for each of the PFTs, as a function of total plant height ($H$).} Each point is an individual plant.}
    \label{fig:figureS2}
\end{figure}

>>>>

The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\cite{ise2010}. Globally, forests store approximately 360Pg of carbon in living biomass\citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}. Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass determines turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}. Reducing uncertainties about biomass partitioning and carbon residence times in GVMs is therefore a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014}.

@@

Moreover, we show that differences in LMF among woody PFTs arise due to differences in leaf morphology, but they converge in the amount of leaf area held on the plant (Fig. 2b-c). Previously, it was unclear whether PFTs with higher leaf biomass also had higher total leaf area. Leaf area is relevant because it drives total light interception and thus photosynthesis. Several classic studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}. We found no evidence for this, instead we found that the three PFTS maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR).

Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig. 2a and S2). We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}. We then found that LMF at a common height was proportional to the leaf construction cost, as measured by leaf mass per area (LMA) (Fig. 2b). As a consequence, LAR was invariant between PFTs, because $\textrm{LAR} = LMF / \textrm{LMA}$ (Fig. 2c and S3).

Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem ($A_S$), a quantity closely related to hydraulic supply capacity \cite{enquist2003}. This ratio is also important because plant leaf mass is frequently estimated from $A_S$ with allometric equations \cite{shinozaki1964b, chave2005}.  Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig. 2d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig. 2e).

The above results highlight how differences in LMA -- a central trait of the leaf economics spectrum\citep{wrightetal2004} -- drives differences in biomass partitioning among dominant woody PFTs. To assess whether LMA could replace PFT when predicting biomass partitioning, we fit a linear mixed-effects model to LMF and $M_F/A_S$ using PFTa dn plant height (and the quadratic term) as predictors (and all interactions). We then replaced PFT with LMA, and found that LMA alone could explain almost as much variation in LMF as PFT ($R^2$ = 0.74 with PFT vs. 0.62 with LMA) and $M_F/A_S$ ($R^2$ = 0.29 vs. 0.28). This is surprising because LMA only captures aspects of leaf morphology. These results suggest promising avenues for parameterizing and simplifying biomass partitioning routines in GVMs.

@@

\emph{I think the point is that we found MF/AS to be proportional to LMA across PFTs, however, the other component of MF/AS is AF/AS and that in fact, across species, explained more variation in MF/AS. So, though LMA is a good substitute for PFT, lots of other variation still exists, which should be taken into account. We just have shown that AF/AS does not depend on PFT, and almost not on climate - so it has to be some unidentified factor instead. ??}. In contrast, we found that variation among species in leaf and stem traits could explain additional variation in biomass partitioning. The effect of LMA was found to hold also within PFTs (Fig. 4a), explaining 28\% of the variation in $M_F/A_S$ across species. The fact that PFTs differ in LMA is evident in the separation of the three groups along the x-axis of Fig. 4a. Yet these clouds were broad and overlapping, with a more general relationship extending across the entire LMA axis. The remaining (?? what is number?? \emph{the remainder should be 100 - 28}) variation in $M_F/A_S$  was explained by species-level differences in $A_F/A_S$ (Fig. 4b). As previously noted, $A_F/A_S$ does not differ systematically among PFTs (Fig. 2e), but does vary close to two orders of magnitude across species. The separation among PFTs along the Y-axis arises due to differences in LMA.

@@

We used the recently compiled Biomass And Allometry Database (BAAD) \citep{Falster2015}, including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca. 1.3m) ($A_{Sbh}$).  For each analysis, we used different subsets of the data. Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4} We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).

@@

We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves). For conifers, leaf area was converted to half-total surface area using published conversion factors \citep{barclay2000}, which is most appropriate for comparison to flat leaves \citep{lang1991, chen1992}. Stem cross-sectional area was measured either at breast height (ca. 1.3m) and/or at the base of the plant. For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using non-linear regression (see~\ref{sec:predas}).

@@

Figure 1a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary. We thank Martin de Kauwe for discussions and comments on an earlier version of this paper.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
    \caption{\textbf{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$).} Each point is an individual plant.}
    \label{fig:figureS2}
\end{figure}
